---
title: "Coursework 2"
subtitle: "MA22019 Introduction to Data Science"
format: gfm
editor: visual
---

<!--- DO NOT DELETE THIS LINE --->

<!--- METADATA ANCHOR: MA22019-CW2-2026 --->

# Data Preparation Notes

<!--- DO NOT DELETE THIS LINE - DATA PREPARATION NOTES ANCHOR --->

Briefly document the main data-handling decisions that matter for your analysis. This section may be short prose or one small table.

```{r}
#| label: data-preparation-code
#| warning: false
#| message: false
# Write your data-preparation code here.

library(tidyverse)
library(sf)
library(ggspatial)
library(prettymapr)
library(patchwork)
library(spatstat)
library(janitor)

city_wards <- read_sf("data/city_wards.shp")
ward_profiles <- read.csv("data/ward_profiles.csv")
incident_reports <- read.csv("data/incident_reports.csv")

ward_profiles <- ward_profiles %>% 
  mutate(ward_name = gsub("-", " ", ward_name)) %>% 
  mutate(rental_share = as.numeric(gsub("%","",rental_share)))

city_wards_profiles <- city_wards %>% 
  left_join(ward_profiles, by = c("ward_id", "ward_name"))

cat("Total wards in city:", nrow(city_wards),"\n")

ward_profiles_missing <- ward_profiles %>% 
  summarise(across(everything(), ~sum(is.na(.))))

incident_sf <- st_as_sf(incident_reports, coords = c("x", "y"), crs = st_crs(city_wards_profiles)) #turn the coords from incident_reports into a sf with the correct coordinate system. then use st_join to join this with the city_wards_profiles outline.

incidents_wards <- st_join(incident_sf, city_wards_profiles)

no_mislabelled_incidents <- incidents_wards %>% 
  filter(ward_name.x != ward_name.y) %>% 
  nrow()    # 39 mislabelled incidents.

cat("No. incident reports with mislabelled wards:", no_mislabelled_incidents, "\n")

incidents_wards <- incidents_wards %>% 
  rename(ward_recorded = ward_name.x, correct_ward = ward_name.y)


```

Write your `Data Preparation Notes` answer here.

I checked the data for inconsistent formatting in any variables. I found that ward_name is inconsistent, as some have hyphens in ward_profiles but not in city_wards. I also found that rental_share in ward_profiles is inconsistent as some values have % signs and some don't. I removed these and asserted that the value was numeric. I found that in ward_profiles there are in 3 missing values in transport_access and 3 missing values in listed_building_share. I have left these in so we don't lose data unnecessarily, but will have to deal with them when doing analysis of these variables. I joined the ward geometries and profiles so that each ward has its info included. I used left_join so that all the info from ward_profiles is added to each ward, without risk of losing any. I joined by ward_id and ward_name as I had already cleaned these so they should match up, and I didn't want columns to duplicate unnecessarily. incident_reports showed no duplicated entries. I double-checked whether the labels for ward lined up with the coordinates given. After converting incident reports to sf and joining to city_wards_profiles I found that 39 entries did not line up. I decided to trust the coordinates given rather than the label, as they are more likely to be accurate, especially for incidents happening on borders which have been mislabelled. 


- ward_name is inconsistently formatted, some have hyphens in ward_profiles
- city_wards and ward_profiles both have 32 unique rows, so this is consistent, i checked all names now match up
- I also checked all the ward_names in incident_reports are formatted the same and match up
- In ward_profiles, rental_share is inconsistently formatted. i have removed % signs and made the values numeric.
- in ward_profiles there are 3 missing values in transport_access and 3 missing values in listed_building_share. These are kept in but will be removed before doing analysis on those variables
- incident_reports and city_wards are both free of missing values
- no duplicates in incident_reports
- i checked consistency between coordinates and labels in incident_reports. i found 39 incidents which had labels which are not consistent with the coordinates given.
- i renamed the ward in which it was recorded as ward_recorded and the ward it lies in by coordinates as correct_ward.
- i am trusting the coordinates to be more accurate as labels may have been put in incorrectly, or the place lies on the border and they were labelled incorrectly
- i joined the ward_profiles to the city_wards using left_join, so that each ward has its relevant info and geometry. i joined by ward_id and ward_name as i already cleaned these to be the same, and i didnt want columns to duplicate


# Task 1

<!--- DO NOT DELETE THIS LINE - TASK 1 ANCHOR --->

Identify and investigate the most important spatial patterns in the incident data.

```{r}
#| label: task-1-code
# Write your Task 1 code here.

incidents_per_ward <- incidents_wards %>% 
  st_drop_geometry() %>% 
  count(correct_ward, name = "n_incidents")

top_incidents_per_ward <- incidents_per_ward %>% 
  arrange(desc(n_incidents)) %>% 
  head(8)

incidents_per_ward_plot <- top_incidents_per_ward %>% 
  ggplot(aes(x = reorder(correct_ward, n_incidents), y = n_incidents)) +
  geom_col(fill = "violet") +
  coord_flip() +
  labs(
    x = "Ward",
    y = "Number of incidents",
    title = "Top 8 wards by incident count") +
  theme_minimal()

spatial_distribution_incidents <- incidents_wards %>% 
  ggplot()+
  geom_sf(data = city_wards, fill = "grey98", colour = "black")+
  geom_sf(data = incidents_wards, alpha = 0.6, size = 0.6, colour = "black")+
  theme_bw()+
  labs(title = "Spatial distribution of incident reports", x = "Longitude", y = "Latitude")

incidents_per_ward_plot + spatial_distribution_incidents

incident_ppp <- ppp(
  st_coordinates(incidents_wards)[,1],
  st_coordinates(incidents_wards)[,2],
  window = as.owin(city_wards))

density_map <- density.ppp(incident_ppp, edge = TRUE, sigma = 3000)
plot(density_map, main = "Kernel-smoothed density of incidents")

areas_counts <- city_wards_profiles %>% 
  left_join(incidents_per_ward, by = c("ward_name" = "correct_ward")) %>% 
  mutate(n_incidents = replace_na(n_incidents, 0))

top_ward <- areas_counts %>% 
  slice_max(n_incidents, n = 1)

choropleth_map <- areas_counts %>% 
  ggplot(aes(fill = n_incidents)) +
    geom_sf() +
    geom_sf_text(data = top_ward, aes(label = ward_name), size = 3, vjust = -1.8, hjust = -0.05)+
    theme_bw() +
    scale_fill_distiller(palette = "Reds", trans = "reverse") +
    labs(
        x = "Longitude", y = "Latitude",
        fill = "Incidents\nper ward",
        title = "Incidents per ward")
choropleth_map
```

Write your Task 1 answer here.

I first grouped by ward and computed the number of incidents per ward. I found that there was significant variation between wards, with some wards such as "Canal Side" having no incidents, and others have many incidents such as "Maple Cross" with 337. I plotted the spatial distribution of incidents, and found that there was a significant North-South divide, with the majority of incidents taking place in the North of the city. More specifically, the North-East was the most incident-heavy region. I plotted the kernel-smoothed incident density to visualise the pattern of expected number of incidents. Again it was clear that most of the incidents occured in the north-east. I plotted a choropleth map to give a better visualisation of which wards suffered the most incidents. Maple Cross was the most significant in terms of incident frequency and, while we also see some hotspots in the east and North-west. 



# Task 2

<!--- DO NOT DELETE THIS LINE - TASK 2 ANCHOR --->

Use the available neighbourhood data to develop and justify possible explanations for the patterns you identified.

```{r}
#| label: task-2-code
# Write your Task 2 code here.
```

Write your Task 2 answer here.

# Task 3

<!--- DO NOT DELETE THIS LINE - TASK 3 ANCHOR --->

Make a small number of evidence-based recommendations for priority areas or actions.

```{r}
#| label: task-3-code
# Write your Task 3 code here.
```

Write your Task 3 answer here.

# References

<!--- DO NOT DELETE THIS LINE - REFERENCES ANCHOR --->

Add references only if needed.

```{r}
#| label: word-count-check
#| echo: false
#| message: false
#| warning: false

resolve_word_count_input <- function() {
  current_input <- tryCatch(
    knitr::current_input(),
    error = function(e) NA_character_
  )

  if (!is.na(current_input) && nzchar(current_input) && file.exists(current_input)) {
    return(current_input)
  }

  if (file.exists("coursework_02.qmd")) {
    return("coursework_02.qmd")
  }

  stop(
    paste(
      "Could not locate coursework_02.qmd.",
      "Render the document from inside the coursework project folder."
    )
  )
}

extract_chunk_captions <- function(chunk_lines) {
  caption_lines <- grep(
    "^#\\|\\s*(fig-cap|tbl-cap|lst-cap)\\s*:",
    chunk_lines,
    value = TRUE
  )

  if (length(caption_lines) == 0) {
    return(character(0))
  }

  trimws(sub("^#\\|\\s*(fig-cap|tbl-cap|lst-cap)\\s*:\\s*", "", caption_lines))
}

count_policy_words <- function(text) {
  if (length(text) == 0) {
    return(0L)
  }

  text <- paste(text, collapse = " ")
  text <- gsub("`[^`]*`", " ", text, perl = TRUE)
  text <- gsub("\\[[^\\]]*\\]\\([^\\)]*\\)", " ", text, perl = TRUE)
  text <- gsub("https?://\\S+", " ", text, perl = TRUE)
  text <- gsub("[*_~#>]", " ", text, perl = TRUE)

  tokens <- regmatches(
    text,
    gregexpr("[[:alnum:]]+(?:[-'][[:alnum:]]+)*", text, perl = TRUE)
  )[[1]]

  if (length(tokens) == 1 && identical(tokens, character(0))) {
    return(0L)
  }

  length(tokens)
}

count_prose_words <- function(file_path) {
  lines <- readLines(file_path, warn = FALSE)

  yaml_bounds <- grep("^---\\s*$", lines)
  if (length(yaml_bounds) >= 2) {
    lines <- lines[-seq.int(yaml_bounds[1], yaml_bounds[2])]
  }

  reference_heading <- grep("^#\\s+References\\s*$", lines)
  if (length(reference_heading) > 0) {
    lines <- lines[seq_len(reference_heading[1] - 1)]
  }

  prose_lines <- character(0)
  caption_lines <- character(0)
  in_chunk <- FALSE
  current_chunk <- character(0)

  for (line in lines) {
    if (grepl("^```", line)) {
      if (!in_chunk) {
        in_chunk <- TRUE
        current_chunk <- character(0)
      } else {
        caption_lines <- c(caption_lines, extract_chunk_captions(current_chunk))
        in_chunk <- FALSE
        current_chunk <- character(0)
      }
      next
    }

    if (in_chunk) {
      current_chunk <- c(current_chunk, line)
      next
    }

    trimmed <- trimws(line)
    if (
      trimmed == "" ||
      grepl("^<!--", trimmed) ||
      grepl("^#+\\s+", trimmed)
    ) {
      next
    }

    prose_lines <- c(prose_lines, line)
  }

  total_words <- count_policy_words(prose_lines) + count_policy_words(caption_lines)
  total_words
}

true_word_count <- count_prose_words(resolve_word_count_input())
word_limit <- 1000L
status_text <- if (true_word_count <= word_limit) "under" else "over"
delta <- abs(word_limit - true_word_count)

cat(
  sprintf(
    "**Prose Word Count:** %d words (%d words %s the %d-word limit)",
    true_word_count,
    delta,
    status_text,
    word_limit
  )
)
```
